---
title: "Continuous Glucose Monitoring (CGM) Data"
output: 
  html_document:
   toc: true
   toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Continuous Glucose Monitoring (CGM) Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction

The `hypometrics` package was developed to manipulate data from CGM devices manufactured by Abbott, specifically the Abbott Freestyle Libre 2 sensor and reader. Future work will involve updating the package so it can function with data from additional CGM devices (e.g. Dexcom).

This article describes the CGM-specific functions that were created as part of the `hypometrics` package. 

### Setup

To be able to use the CGM functions, firstly install and load `hypometrics`.

```
#Install
install.packages("remotes")
remotes::install_github("hypometrics")
```

```{r, message=FALSE}
#Load package
library(hypometrics)
```

### Example data

Throughout this tutorial, the examples presented will be based on the [`raw_cgm`](https://hypometrics.github.io/reference/raw_cgm.html) and [`cgm`](https://hypometrics.github.io/reference/cgm.html) dataframes.  

Both dataframes contain data from `r dplyr::n_distinct(cgm$id)` participants who used a CGM device, with one row per CGM timestamp. The `cgm` dataframe is a cleaner and more granular version of the `raw_cgm` as it includes glucose data that already been interpolated, at the 1-min level. Further information on data interpolation is provided in the next section. 

* A preview of the `cgm` dataframe is shown below:

```{r, message=FALSE}
utils::head(cgm)
```

## 2. Cleaning CGM data

### From implicit to explicit gaps in CGM data

The [`raw_cgm`](https://hypometrics.github.io/reference/raw_cgm.html) dataset has gaps in the CGM timestamps which are implicit. For example, as shown below, there is a gap between 17:00 and 17:20 but the missing CGM timestamps and glucose values are not explicitly included in the dataset. This means that gaps in the data are not particularly obvious to the user. 

```{r, message=FALSE}
raw_cgm %>% dplyr::slice(26:36)
```

\

The `cgmInterpolate()` function addresses this by turning these implicit gaps into explicit gaps. In that case, it will add 17:05, 17:10 and 17:15 as timestamps with the corresponding glucose values marked as missing (i.e. NA), as shown below. The 'Interpolate' argument is set to FALSE as the aim here is solely to produce explicit gaps, not to interpolate glucose values. The 'MinGap' argument determines the minimum data gap considered and it is set to 600 seconds (or 10 min) so at least 1 CGM timestamp can be added. The desired granularity of data is 5 minutes so the 'Granularity' argument is set to 300 seconds.

\

```{r, message=FALSE}
cgm_explicit_gaps <- cgmInterpolate(raw_cgm, Interpolate = FALSE, MinGap = 600, Granularity = 300)

cgm_explicit_gaps %>% dplyr::slice(31:41)
```

\

### Interpolation of CGM data

There might be instances where, following turning implicit gaps into explicit gaps, missing glucose values need to be interpolated. This can be done by simply changing the 'Interpolate' argument of the same function to TRUE. This will allow the linear interpolation of glucose data using the `stats::approx()` function. The 'MaxGap' argument here becomes essential as it determines what is the maximum gap for which glucose values should be interpolated. For example, if set at 1800 seconds (the default in the function), any glucose values in a gap longer than 30 minutes will not be interpolated and left as NA. It is up to the user to determine what the 'MaxGap' should be depending on their analysis. Let's see what the linear interpolation looks like below, still based on our 17:00 to 17:20 gap example from earlier.


```{r, message=FALSE}
cgm_explicit_gaps_and_interpolation <- cgmInterpolate(raw_cgm, Interpolate = TRUE, MinGap = 600, Granularity = 300, MaxGap = 1800)

cgm_explicit_gaps_and_interpolation %>% dplyr::slice(31:41)
```


We can see here that as the gap was less than 30 minutes, glucose values have been linearly interpolated. The added values are 3.28 mmol/L (at 17:05), 3.15 mmol/L (at 17:10) and 3.02 mmol/L (at 17:15).

## 3. Checking CGM data

After cleaning the data and before diving into data analysis, the `checkMissingCGM()` function enables the evaluation of missingness in glucose data. 

### Sample level

Firstly, you can explore the number of hours of CGM data available per day for all participants included by running: 

```{r, message=FALSE}
cgmCheck(cgm)
```

### Individual level

With the initial identification of days/participants with low CGM data coverage in the previous step, further investigation can be conducted by looking at CGM data from individual participants. For example:

```{r, message=FALSE}
missingness_001 <- cgmCheck(cgm, CheckAll = FALSE, StudyID = "001", AxisLabels = c(0, 2.2, 3.9, 10, 20))
```
This creates a list which contains a data set with daily breakdown of available hours of CGM data for participant '001' and a graph highlighting periods of missingness as shown below. 

```{r, message=FALSE, warning=FALSE}
graphics::plot(missingness_001$cgm_na_distribution_plot)
```

## 4. Describing CGM data

### Continuous glucose monitoring data summary

Using the `cgmSummarise()` function, which combines the outputs of multiple [iglu](https://irinagain.github.io/iglu/index.html) package functions, a summary of CGM data can be produced. This includes CGM data coverage, mean/median glucose, glucose standard deviation/interquartile range, coefficient of variation, time above, below and in range for each participant. The default unit for glucose values is mmol/L but this can be changed to mg/dL. The parameters for the interquartile range, time in/above/below range can also be adapted depending on user's needs. 

```{r, message=FALSE, warning=FALSE}
cgmSummarise(DataFrame = cgm, GlucoseUnit = "mmol/L")
```

### Hypoglycaemia episode detection

The `sdhDetection()` function allows to automatically detect and report on every episode of CGM-detected hypoglycaemia. The function takes CGM data and returns for each participant all episodes of hypoglycaemia that were detected: start and end time of episode, duration, nadir and day/night status (night: 00h00-06h00). If an episode overlaps between day and night status, it is marked as overlap. The default hypoglycaemia detection limit is 3.9 mmol/L and duration 15 minutes.

```{r, message=FALSE}
utils::head(sdhDetection(DataFrame = cgm), n=10)
```

The default parameters can be changed to explore a different glucose threshold, and minimum duration. For example: 

```{r, message=FALSE}
utils::head(sdhDetection(DataFrame = cgm, DetectionLimit = 3, DetectionDuration = 30), n=10)
```


If sleep data is available, the user can also sleep information to the output by changing the default AddSleepStatus parameter to "yes". For this to run, the CGM dataset used must include a sleep_status column (this can be obtained using the `cgmsleepLink()` function). For example:

```{r, message=FALSE}
linked_data <- cgmsleepLink(CgmDataFrame = hypometrics::cgm,
                            SleepDataFrame = hypometrics::raw_sleep)

utils::head(sdhDetection(DataFrame = linked_data), n=10)
```

### Sensor-detected hypoglycaemia data summary

Once the list and details of all episodes of hypoglycaemia have been produced, it might be useful to have summarised descriptive data in a single dataset. This can be done using the `sdhSummarise()` function. With one row per participant, the output includes the overall number of hypoglycaemic episodes, the number occuring the night or day, the mean duration of those episodes as well as details on longer episodes of hypoglycaemia. The user can define how a long episode is defined, using the 'LongDuration' argument. The default is 120 minutes. This summary can be produced at any threshold of glucose, as it simply takes the output of the `sdhDetection()` function as input. Let's see a summary of SDH data, for a glucose threshold of 3.9 mmol/L (the default), using our example dataset.

```{r, message=FALSE}
## Producing the list of hypoglycaemia episodes for all participants
sdh_map <- sdhDetection(DataFrame = cgm)

## Summarising SDH data
sdhSummarise(DataFrame = sdh_map)
```

If sleep data is available, the parameter `AddSleepSummary` can be turned to "yes" and the number of SD when asleep, awake or when sleep information was missing for each SDH and participant will also be calculated.

## 5. Visualising CGM data

It may be useful to plot CGM data to visually inspect glucose concentrations over time. The `cgmVisualise()` function allows you to do this at three different levels of granularity. 

### Overall

Using the default function parameters, you will obtain an overview of glucose data for the entire study period for a selected participant, as shown below:

```{r, message=FALSE, warning=FALSE}
cgmVisualise(cgm, StudyID = "001")
```

The figure shows the glucose trace in blue, the 3.9-10mmol/L range is delimited by green lines. The grey area highlights the period from 00:00 to 06:00 for reference, as it is typically used when examining nocturnal hypoglycaemia.

**Note:** if sleep tracker data is available, the user can plot this instead by specifying "yes" to the `AddSleep` argument of the function (default is "no") and the `DataFrame` used must be one where CGM and sleep data are linked.

### Week by week

It is also possible to view glucose data on a weekly basis by specifying a breakdown by week in the `TimeBreak` argument of the function and which week is of interest using the `PageNumber` argument, as shown below:

```{r, message=FALSE, warning=FALSE}
cgmVisualise(cgm, StudyID = "001", TimeBreak = "week", PageNumber = 3)
```
\

The number at the top of the figure indicates the week selected for visualisation. Please note the function will return an error if the picked PageNumber (i.e. here, week number) is out of the data range (e.g. PageNumber = 11 when there are only 10 weeks of data).

### Day by day

Lastly, for further granularity, there is an option to visualise this data for specific days using the same logic as for the weekly data. 

```{r, message=FALSE, warning=FALSE}
cgmVisualise(cgm, StudyID = "001", TimeBreak = "day", PageNumber = 2)
```
\

The number at the top of the figure indicates the day selected for visualisation. Please note the function will return an error if the picked PageNumber (i.e. here, day number) is out of the data range (e.g. PageNumber = 71 when there are only 70 days of data).


