---
title: "Continuous Glucose Monitoring (CGM) Data"
output: rmarkdown::html_vignette
---

---
vignette: >
  %\VignetteIndexEntry{Continuous Glucose Monitoring (CGM) Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction

The `hypometrics` package was developed to manipulate data from CGM devices manufactured by Abbott, specifically the Abbott Freestyle Libre 2 sensor and reader. Future work will involve updating the package so it can function with data from additional CGM devices (e.g. Dexcom).

This article describes the CGM-specific functions that were created as part of the `hypometrics` package. 

### Setup

To be able to use the CGM functions, firstly install and load `hypometrics`.

```
#Install
install.packages("remotes")
remotes::install_github("hypometrics")
```

```{r, message=FALSE}
#Load package
library(hypometrics)
```

### Example data

Throughout this tutorial, the examples presented will be based on the [`cgm`](https://hypometrics.github.io/reference/cgm.html) dataframe.  

* This dataframe contains data from `r dplyr::n_distinct(cgm$id)` participants who used a CGM device, with one row per CGM timestamp as shown below.

```{r, message=FALSE}
utils::head(cgm)
```

## 2. Cleaning CGM data

### Interpolation

## 3. Analysing CGM data

### Hypoglycaemia episode detection

The `sdhDetection()` function allows to automatically detect and report on every episode of CGM-detected hypoglycaemia. The function takes CGM data and returns a list of all episodes of hypoglycaemia including their characteristics such as duration and nadir. The default hypoglycaemia detection limit is 3.9 mmol/L and duration 15 minutes. 

## 4. Checking

## 5. Visualising








