---
title: "Sleep Data"
output: 
  html_document:
   toc: true
   toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Sleep Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction

The `hypometrics` package was built to handle physical activity and sleep data generated by the FitBit Charge 4. Future work will consist of augmenting the package's adaptability in order to read in data from different Fitbit devices or other activity and sleep trackers (e.g. Garmin).

This article describes the sleep-specific functions that were created as part of the `hypometrics` package.


### Setup

To be able to use the sleep functions, firstly install and load `hypometrics`.

```
#Install
install.packages("remotes")
remotes::install_github("leicester-cdag/hypometrics")
```


```{r, message=FALSE}
#Load package
library(hypometrics)
```

### Simulated data

Throughout this tutorial, the examples presented will be based on the [`raw_sleep`](https://leicester-cdag.github.io/hypometrics/reference/raw_sleep.html) simulated dataset. 

* A preview of the `raw_sleep` dataframe is shown below:

```{r, message=FALSE}
utils::head(raw_sleep)
```

## 2. Reading sleep data 

The function `sleepRead()` allows the user to read and combine raw sleep files downloaded directly from their Fitbit account. If the downloaded folder is zipped, there is the option to unzip the folder by setting the `Unzip` argument to TRUE. This will create a new folder in the selected `Folder Path` called `Unzipped Fitbit`. The Fitbit file type to read must be specified using the `FileType` argument, either json or csv. As the individual files do not contain personal identifiers, it might be useful to add an ID column across the files to easily identify participants. It is possible to do this using the `StudyID` argument of the function.

An example of the syntax is shown below: 
```{r, eval = FALSE}
hypometrics::sleepRead(Unzip = F,
                       FolderPath = "C:/Users",
                       FileType = "json",
                       StudyID = "P02")
```

## 3. Cleaning sleep data

The `sleepCategorise()` function allows the user to separate the raw sleep data into three distinct datasets that can be used for specific needs. 

### Overview of sleep periods  

The `sleep_overview` is a dataset containing overall sleep information including sleep start and end times as well as the number of minutes awake, time spent in bed and duration of sleep as shown below. Each sleep event is tagged with a specific logId (internal to Fitbit). The logId can be used to identify specific sleep events and merge sleep information across the three datasets. The first 8 columns of the `sleep_overview` dataset are shown below.

```{r, message=FALSE, warning=FALSE}
sleepCategorise(raw_sleep) %>%
  dplyr::slice(1:6)
```

## 4. Describing sleep data

The `sleepSummarise` function allows the user to summarise raw sleep data to obtain a dataset with one row per participant including the number of nights with missing data, average time in bed, asleep or awake.

```{r, message=FALSE}
sleepSummarise(raw_sleep)
```

## 5. Visualising sleep data

The `sleepVisualise` function allows the user to visualise the distribution of times when participants went to bed and got out of bed, as shown below. The default is to plot the histogram for all times recorded by the Fitbit for all participants. To facilitate interpretation of the sleeping patterns, the function centres the times when going to bed around midnight.

```{r, message=FALSE, warning = FALSE, eval = FALSE}
sleepVisualise(raw_sleep)
```

If the user needs to visualise data for a specific participant, this can be done by changing the argument `VisualiseAll` to FALSE and indicating the participant id of interest in the `StudyID` argument, as shown below:

```{r, message=FALSE, warning = FALSE}
sleepVisualise(raw_sleep, VisualiseAll = FALSE, StudyID = "P02")
```





